---
title: "Adjust samples to match timpoint+infection for each donor"
output: html_notebook
---
Author: Tyler Hansen  
Date: 5.15.24

In order to model the data properly, we need to look at the samples and see how to match timepoint and infection for each donor.

## Setup 

Packages
```{r, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
```

Load in cts and metadata
```{r}
cts <- readRDS("data/filtered_counts_QCd_samples_batch1.rds")
sample_info <- readRDS("data/metadata_with-covariates_QCd_samples_batch1.rds")
```

## Are any rows duplicates?
```{r}
# Show rows that are duplicates
sample_info[duplicated(sample_info), ] %>% nrow()
```

Sweet no duplicates. 

## Make a heatmap of batch1 samples
Facet by infection, x = donor_id, y = timepoint, value = number of samples

```{r, fig.width=6, fig.height=4}
#make matrix of values to plot:
si_subset <- dplyr::select(sample_info , Donor_ID, Timepoint, Infection)

#reverse order of timepoint
si_subset$Timepoint <- factor(si_subset$Timepoint, levels = rev(levels(si_subset$Timepoint)))

#count number of samples, regardless of run
si_subset_count <- count(si_subset, Donor_ID, Timepoint, Infection)

# Convert n to a factor
si_subset_count$n <- factor(si_subset_count$n)

heatmap_all_samples <- ggplot() +
  geom_tile(data = si_subset, aes(x=Donor_ID, y=Timepoint, fill = Infection), color = "white", linewidth = 0.1) +
  facet_wrap(~Infection) +
  ggtitle("RNA-seq samples after QC") +
  xlab("Individual") +
  ylab("Timepoint (hours)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 1,
        text = element_text(colour = "black"), legend.position = "none", 
        strip.background = element_rect(fill = "white", color = "grey"), strip.text = element_text(color = "black")) +
  scale_fill_manual(values = c("snow3", "slateblue"), na.value = 'grey')
heatmap_all_samples

ggsave(heatmap_all_samples, filename = "results/batch1_only/QCd_sample_heatmap_preDuplication.pdf", device = cairo_pdf, width = 6, height = 4)
```

We face two moderate problems:
1) Not all timepoints have NI samples. 
2) Some samples are missing. 

## Duplicate NI samples
Since our downstream analyses are timepoint by timepoint, we will address issue 1 by duplicating NI to the nearest Mtb timepoints. 

Specifically we will do this:
  T0 --> T1/T2
  T4 --> T3/T5/T6
  T8 --> T7/T9
  T12 --> T10/T14
  T18 --> T16
  T48 --> T42

Function for duplicating timepoints
```{r}
duplicate_timepoints <- function(si, reference_t, assigned_t, counts) {
  #extract the desired timepoint
  tmp <- dplyr::filter(si, Infection == "NI" & Timepoint == reference_t)
  
  #change to new timepoint
  tmp <- mutate(tmp, Timepoint = assigned_t)
  
  #change rownames so there are no dupliicates
  rnames <- rownames(tmp)
  rownames(tmp) <- paste0("duplicated_", rnames)
  
  #add back to si
  si <- bind_rows(si, tmp)
  
  #now adjust counts with for loop over rnames. Duplicate sample with rname and add duplicated to it. 
  counts <- as.data.frame(counts)
  for (sample in rnames) {
    dup_sample <- paste0("duplicated_", sample)
    counts[,dup_sample] <- counts[, sample]
    rm(dup_sample)
  }
  
  #this should work, but stop if si and counts are not equivalent. If 
  if (all(colnames(counts) == rownames(si))) {
    res <- list(counts, si)
    return(res)
  }
  else {
    stop("Sample info and counts are not equivalent after operation")
  }
}
```

T0 --> T1
```{r}
adjusted <- duplicate_timepoints(si = sample_info , reference_t = "T0", assigned_t = "T1", counts = cts)
```

T0 --> T1 --> T2
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T1", assigned_t = "T2", counts = adjusted[[1]])
```

T4 --> T3
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T4", assigned_t = "T3", counts = adjusted[[1]])
```

T4 --> T3 --> T5
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T3", assigned_t = "T5", counts = adjusted[[1]])
```

T4 --> T3 --> T5 --> T6
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T5", assigned_t = "T6", counts = adjusted[[1]])
```

T8 --> T7
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T8", assigned_t = "T7", counts = adjusted[[1]])
```

T8 --> T7 --> T9
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T7", assigned_t = "T9", counts = adjusted[[1]])
```

T12 --> T10
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T12", assigned_t = "T10", counts = adjusted[[1]])
```

T12 --> T10 --> T14
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T10", assigned_t = "T14", counts = adjusted[[1]])
```

T18 --> T16
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T18", assigned_t = "T16", counts = adjusted[[1]])
```

T48 --> T42
```{r}
adjusted <- duplicate_timepoints(si = adjusted[[2]], reference_t = "T48", assigned_t = "T42", counts = adjusted[[1]])
```

Extract cts and si.
```{r}
cts <- adjusted[[1]]
sample_info  <- adjusted[[2]]
```

Plot new heatmap. 
```{r, fig.width=6, fig.height=4}
#prepare for heatmap
si_subset <- dplyr::select(sample_info , Donor_ID, Timepoint, Infection)
si_subset$Timepoint <- factor(si_subset$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))
si_subset_count <- count(si_subset, Donor_ID, Timepoint, Infection)
si_subset_count$n <- factor(si_subset_count$n)

#heatmap
heatmap_duplicated_samples <- ggplot() +
  geom_tile(data = si_subset, aes(x=Donor_ID, y=Timepoint, fill = Infection), color = "white", linewidth = 0.1) +
  facet_wrap(~Infection) +
  ggtitle("RNA-seq samples after QC and NI duplication") +
  xlab("Individual") +
  ylab("Timepoint (hours)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 1,
        text = element_text(colour = "black"), legend.position = "none", 
        strip.background = element_rect(fill = "white", color = "grey"), strip.text = element_text(color = "black")) +
  scale_fill_manual(values = c("snow3", "slateblue"), na.value = 'grey')
heatmap_duplicated_samples

ggsave(heatmap_duplicated_samples, filename = "results/batch1_only/QCd_sample_heatmap_NI_Duplicated.pdf", device = cairo_pdf, width = 6, height = 4)
```

This looks great. We now have the more difficult issue of missing data. 

## Duplicate the nearest timepoint for missing NI samples

I assume that NI samples should not vary much across time. We can therefore duplicate the nearest NI timepoint within the donor. 

AF95/NI: T14 --> T16
AF95/NI: T14 --> T16 --> T18
EU144/NI: T6 --> T2 #we want T6 because this was the last point that was duplicated and using T3 or T4 would have duplicate rownames. In other words, T6 is T4. 
EU144/NI: T6 --> T2 --> T1
EU144/NI: T6 --> T2 --> T1 --> T0


```{r}
duplicate_timepoints_2 <- function(si, reference_t, assigned_t, counts, condition, donor) {
  #extract the desired timepoint
  tmp <- dplyr::filter(si, Infection == condition & Timepoint == reference_t & Donor_ID == donor)
  
  #change to new timepoint
  tmp <- mutate(tmp, Timepoint = assigned_t)
  
  #change rownames so there are no dupliicates
  rnames <- rownames(tmp)
  rownames(tmp) <- paste0("duplicated_", rnames)
  
  #add back to si
  si <- bind_rows(si, tmp)
  
  #now adjust counts with for loop over rnames. Duplicate sample with rname and add duplicated to it. 
  counts <- as.data.frame(counts)
  for (sample in rnames) {
    dup_sample <- paste0("duplicated_", sample)
    counts[,dup_sample] <- counts[, sample]
    rm(dup_sample)
  }
  #this should work, but stop if si and counts are not equivalent. If 
  if (all(colnames(counts) == rownames(si))) {
    res <- list(counts, si)
    return(res)
  }
  else {
    stop("Sample info and counts are not equivalent after operation")
  }
}
```

```{r}
#AF95/NI: T14 --> T16
adjusted_2 <- duplicate_timepoints_2(si = sample_info , counts = cts, reference_t = "T14", assigned_t = "T16", condition = "NI", donor = "AF95")

#AF95/NI: T14 --> T16 --> T18
adjusted_2 <- duplicate_timepoints_2(si = adjusted_2[[2]], counts = adjusted_2[[1]], reference_t = "T16", assigned_t = "T18", condition = "NI", donor = "AF95")

#EU144/NI: T6 --> T2
adjusted_2 <- duplicate_timepoints_2(si = adjusted_2[[2]], counts = adjusted_2[[1]], reference_t = "T6", assigned_t = "T2", condition = "NI", donor = "EU144")

#EU144/NI: T6 --> T2 --> T1
adjusted_2 <- duplicate_timepoints_2(si = adjusted_2[[2]], counts = adjusted_2[[1]], reference_t = "T2", assigned_t = "T1", condition = "NI", donor = "EU144")

#EU144/NI: T6 --> T2 --> T1 --> T0
adjusted_2 <- duplicate_timepoints_2(si = adjusted_2[[2]], counts = adjusted_2[[1]], reference_t = "T1", assigned_t = "T0", condition = "NI", donor = "EU144")
```

Extract cts and si. Plot new heatmap. 
```{r}
cts <- adjusted_2[[1]]
sample_info  <- adjusted_2[[2]]
```

```{r, fig.width=6, fig.height=4}
#prepare for heatmap
si_subset <- dplyr::select(sample_info , Donor_ID, Timepoint, Infection)
si_subset$Timepoint <- factor(si_subset$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))
si_subset_count <- count(si_subset, Donor_ID, Timepoint, Infection)
si_subset_count$n <- factor(si_subset_count$n)

#heatmap
heatmap_NI_filled_samples <- ggplot() +
  geom_tile(data = si_subset, aes(x=Donor_ID, y=Timepoint, fill = Infection), color = "white", linewidth = 0.1) +
  facet_wrap(~Infection) +
  ggtitle("RNA-seq samples after QC and NI duplication") +
  xlab("Individual") +
  ylab("Timepoint (hours)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 1,
        text = element_text(colour = "black"), legend.position = "none", 
        strip.background = element_rect(fill = "white", color = "grey"), strip.text = element_text(color = "black")) +
  scale_fill_manual(values = c("snow3", "slateblue"), na.value = 'grey')
heatmap_NI_filled_samples

ggsave(heatmap_NI_filled_samples, filename = "results/batch1_only/QCd_sample_heatmap_NI_Duplicated_NI-filled.pdf", device = cairo_pdf, width = 6, height = 4)
```

## Remove the matching NI condition/donor for missing Mtb samples

Unlike NI, I think it is a poor assumption to assume that Mtb samples will not vary much across time. Therefore, I think it maybe misleading if we were to duplicate the Mtb sample. 

Instead, lets remove the matching NI sample for the two missing points. This should not effect modeling for timepoint-specific ancestry or infection effects, other than reducing power a bit for timepoints T14 and T18. 

Remove: 
AF193/NI/T14
EU144/NI/T18

```{r}
sample_info <- dplyr::filter(sample_info, !(Infection == "NI" & Timepoint == "T14" & Donor_ID == "AF193")) %>%
  dplyr::filter(!(Infection == "NI" & Timepoint == "T18" & Donor_ID == "EU144"))
```

Square with cts. Plot new heatmap. 
```{r}
cts <- cts[,rownames(sample_info)]
```

```{r, fig.width=6, fig.height=4}
#prepare for heatmap
si_subset <- dplyr::select(sample_info , Donor_ID, Timepoint, Infection)
si_subset$Timepoint <- factor(si_subset$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))
si_subset_count <- count(si_subset, Donor_ID, Timepoint, Infection)
si_subset_count$n <- factor(si_subset_count$n)

#heatmap
heatmap_final_samples_T0 <- ggplot() +
  geom_tile(data = si_subset, aes(x=Donor_ID, y=Timepoint, fill = Infection), color = "white", linewidth = 0.1) +
  facet_wrap(~Infection) +
  ggtitle("RNA-seq samples after QC and NI duplication") +
  xlab("Individual") +
  ylab("Timepoint (hours)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 1,
        text = element_text(colour = "black"), legend.position = "none", 
        strip.background = element_rect(fill = "white", color = "grey"), strip.text = element_text(color = "black")) +
  scale_fill_manual(values = c("snow3", "slateblue"), na.value = 'grey')
heatmap_final_samples_T0

ggsave(heatmap_final_samples_T0, filename = "results/batch1_only/QCd_sample_heatmap_final_samples_withT0.pdf", device = cairo_pdf, width = 6, height = 4)
```

## Remove NI T0 samples and save both versions
When modeling we will not need T0 and it could be confounding, so remove that timepoint. It may be useful to plot though, so lets keep it as and RDS. 

Reset timepoint_hr column
```{r}
sample_info  <- mutate(sample_info, Timepoint_hr = as.numeric(str_remove(Timepoint, "T"))) %>%
  relocate(Timepoint_hr, .after = Timepoint)
```

Save with T0
```{r}
saveRDS(sample_info , file = "data/filtered_counts_final_batch1_withT0.rds")
saveRDS(cts, file = "data/metadata_final_batch1_withT0.rds")
```

Remove T0
```{r}
sample_info <- dplyr::filter(sample_info , Timepoint != "T0")
cts <- cts[, row.names(sample_info )]
```

Plot
```{r, fig.width=6, fig.height=4}
#prepare for heatmap
si_subset <- dplyr::select(sample_info , Donor_ID, Timepoint, Infection)
si_subset$Timepoint <- factor(si_subset$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))
si_subset_count <- count(si_subset, Donor_ID, Timepoint, Infection)
si_subset_count$n <- factor(si_subset_count$n)

#heatmap
heatmap_final_samples <- ggplot() +
  geom_tile(data = si_subset, aes(x=Donor_ID, y=Timepoint, fill = Infection), color = "white", linewidth = 0.1) +
  facet_wrap(~Infection) +
  ggtitle("RNA-seq samples after QC and NI duplication") +
  xlab("Individual") +
  ylab("Timepoint (hours)") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 1,
        text = element_text(colour = "black"), legend.position = "none", 
        strip.background = element_rect(fill = "white", color = "grey"), strip.text = element_text(color = "black")) +
  scale_fill_manual(values = c("snow3", "slateblue"), na.value = 'grey')
heatmap_final_samples

ggsave(heatmap_final_samples, filename = "results/batch1_only/QCd_sample_heatmap_final_samples.pdf", device = cairo_pdf, width = 6, height = 4)
```

Adjust factor levels before saving. 
```{r}
#before, adjust factors to remove levels that were removed
sample_info$flow_cell <- factor(as.character(sample_info$flow_cell))
sample_info$Donor_ID <- factor(as.character(sample_info$Donor_ID))
sample_info$Timepoint <- factor(sample_info$Timepoint, levels = c("T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48"))
```

Save without T0
```{r}
saveRDS(cts, file = "data/filtered_counts_final_batch1.rds")
saveRDS(sample_info, file = "data/metadata_final_batch1.rds")
```