---
title: "Remove Poor Quality Samples and Filter Counts"
output:
  html_document:
    df_print: paged
---
Author: Tyler Hansen  
Date: 3.15.24

I prepared a counts matrix and a metadata table. Here I will investigate them and remove samples that are poor quality. Prior to removing, I will first filter counts so that only expressed protein coding genes are analyzed in subsequent analysis. 

We will evaluate this in the following way:
1) Determine if there are any library count size outliers.
2) Determine if the count distribution for each sample is roughly the same. 
3) Determine if any samples have NA values. 

## Setup 

Packages
```{r, echo=FALSE}
suppressPackageStartupMessages(library(tidyverse))
library(limma)
library(edgeR)
```

Load in the "three tables" - cts|metadata|genes
```{r}
cts <- readRDS("data/kallisto.gene-level.lengthScaledTPM.rds")
sample_info <- readRDS("data/metadata_with-covariates.rds")
gtf <- rtracklayer::readGFFAsGRanges("data/Homo_sapiens.GRCh38.87.gtf")
```

### Filter Counts

1. Extract only the protein coding genes
```{r}
#extract protein coding genes. I checked and this entire pipe works. 
pc_gene_ids <- plyranges::filter(gtf, type == 'gene', gene_biotype == 'protein_coding') %>% as.data.frame() %>% dplyr::select(gene_id, gene_name)

#select rows that match the protein coding gene
filtered_cts_pc_only <- cts[pc_gene_ids$gene_id, ]

#verify this worked - both of these should print TRUE
all(rownames(filtered_cts_pc_only) == pc_gene_ids$gene_id)
nrow(filtered_cts_pc_only) == nrow(pc_gene_ids)
```

Since we have the symbol and ensembl ids now, merge them together and set as rownames. 
```{r}
pc_gene_ids[, 3] <- unite(pc_gene_ids, "merged_id", gene_id:gene_name, sep = "_")
rownames(filtered_cts_pc_only) <- pc_gene_ids$merged_id
```

2. remove low count data by voom-transformed row medians > 1.5
```{r}
#set up dgelist
tmp <- DGEList(filtered_cts_pc_only)
#calculate norm factors
tmp <- calcNormFactors(tmp)
#voom transform, this requires a design, which can be ~1. This gives the same genes below if a real model is used (I checked). 
tmp_design <- model.matrix(~1, data = sample_info)
tmp <- voom(tmp, tmp_design, plot = F)

#filter genes based on voom-transformed row medians > 1.5. This is equivalent to 10^1.5 = 31.6 raw counts. This is much more stringent.  
keep <- data.frame(medians = apply(tmp$E, 1, median)) %>% dplyr::filter(medians > 1.5)
filtered_cts <- filtered_cts_pc_only[row.names(keep), ]

#remove tmp objects
rm(tmp, tmp_design)

#report how much was lost
print('total features remaining')
nrow(filtered_cts)
print('total features removed') 
nrow(filtered_cts_pc_only) - nrow(filtered_cts)
```

## Identify Bad Samples 

### Library size outliers

Set up dge_list with norm factors
```{r}
#set up dgelist
dge <- DGEList(filtered_cts)

#calculate norm factors
dge <- calcNormFactors(dge)
```

Plot library size
```{r}
lib <- dge$samples %>% as.data.frame()
lib_text <- rownames_to_column(lib) %>% dplyr::top_n(n = -2, wt = lib.size)

ggplot(data=lib, aes(x=log10(lib.size))) +
    geom_histogram(binwidth = 0.05, color = "grey", fill = "ivory") +
    geom_jitter(color = "black", alpha = 0.5, aes(y = 0), height = 5) +
    geom_text(data = lib_text, aes(x=log10(lib.size), label = rowname, y = 1), size = 3, angle = 90, hjust = -0.1) +
    theme_bw() +
    ylab("count") +
    theme(legend.position = "none")
```

We have two library count outliers. Let's remove these samples. 
```{r}
bad_samples <- lib_text$rowname
filtered_cts_goodSamples <- filtered_cts[, -which(colnames(filtered_cts) %in% bad_samples)]

#verify this worked
ncol(filtered_cts) - ncol(filtered_cts_goodSamples)
```

Remove them from the sample_info file. 
```{r}
sample_info_goodSamples <- sample_info[-which(rownames(sample_info) %in% bad_samples), ]

#verify this worked
nrow(sample_info) - nrow(sample_info_goodSamples)
```
Check that correct samples were removed by plotting again: 
```{r}
#set up dgelist
dge_good <- DGEList(filtered_cts_goodSamples)

#calculate norm factors
dge_good <- calcNormFactors(dge_good)

```

```{r}
lib_good <- dge_good$samples %>% as.data.frame()

ggplot(data=lib_good, aes(x=log10(lib.size))) +
    geom_histogram(binwidth = 0.05, color = "grey", fill = "ivory") +
    geom_jitter(color = "black", alpha = 0.5, aes(y = 0), height = 5) +
    theme_bw() +
    ylab("count") +
    theme(legend.position = "none")
```

Great, the library sizes look good. Move onto the next QC check: count distribution. 

### Verify count distribbution is good  

Voom Transform
```{r}
#voom transform, this requires a design, which can be ~1 since we are only doing voom transformations, not modeling yet.  
design_base <- model.matrix(~1, data = sample_info_goodSamples)
v <- voom(dge_good, design_base, plot = F)
```

Create df of voom counts with gene, sample, and counts as columns
```{r}
voom_counts_df <- as.data.frame(v$E) %>% rownames_to_column("gene_id") %>% 
  pivot_longer(cols = -gene_id, names_to = "Sample", values_to = "counts")
```

Plot
```{r}
vc_dist <- ggplot(voom_counts_df, aes(x=counts, group=Sample, color=Sample)) +
    geom_density(adjust=1.5, alpha=.4) +
    theme_minimal() +
    theme(legend.position = "none")

vc_dist
```

Great, they are all nicely normalized and comparable across samples. We do not have to remove samples from this QC step. 

### Check that there are no NAs for any gene or samples.
```{r}
#this prints the number of samples that have NAs
which(is.na(filtered_cts_goodSamples))
```
This worked, there are no NA values. We do not have to remove samples from this QC step. 

## Write out the good filtered counts and sample_info
```{r}
#cts
saveRDS(filtered_cts_goodSamples, file = "data/filtered_counts_QCd_samples.rds")

#sample info
saveRDS(sample_info_goodSamples, file = "data/metadata_with-covariates_QCd_samples.rds")
```

