---
title: "Generate Metadata Heatmaps"
output:
  html_document:
    df_print: paged
---
Author: Tyler Hansen  
Date: 11.17.2023 | Update 12.7.2023

From processes I did in bash, I prepared a tsv of metadata information from file names. I will load in the tsv to summarise and build visuals of the data structure. 
```{r, libraries, message=FALSE}
library(tidyverse)
library(extrafont)
loadfonts(device = "win")
```

```{r}
CN <- c("kallisto_number", "Donor_ID", "Timepoint", "Infection(TB/NI)", "MOI", "Location", "Run", "filename", "sample_ID") #column names
md <- read_tsv("results/metadata.tsv", col_names = CN)

head(md)
```
Okay, I checked the md tibble and everything read in properly. There are 668 rows, i.e. 668 samples to analyze. 

First question is are any rows duplicates?
```{r}
# Check for duplicate rows
duplicate_rows <- md %>% duplicated()
 
# Show rows that are duplicates
md[duplicate_rows, ] 

```

Sweet no duplicates. 

Now plot a heatmap matrix of individuals by timepoint and the number of samples for that. Like what Haley had done before. 

```{r}
#make matrix of values to plot:
tp_donor_count <- count(md, Donor_ID, Timepoint)

#adjust factor levels so timepoints are in order (donors are pretty good already):
tp_donor_count$Timepoint <- factor(tp_donor_count$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))
tp_donor_count
```

```{r}
# Convert n to a factor
tp_donor_count$n <- factor(tp_donor_count$n)

#plot as heatmap:
p <- ggplot(tp_donor_count, x=Donor_ID) +
  geom_tile(aes(x=Donor_ID, y=Timepoint, fill = n), colour="white", size=0.2) +
  geom_text(aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  labs(fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), legend.position = "none", text = element_text(family = "Calibri", colour = "black")) +
  scale_fill_brewer(palette = "Blues")
p

#save
ggsave(p, device = cairo_pdf, filename = "results/metadata_heatmap_timepoint-vs-donor.pdf")
```
Okay so we know the overall stucture, but it isn't clear whether 2 is NI and MtB or NI and Mtb. Expand to include additional info.  

```{r}
#make matrix of values to plot:
tp_four <- data.frame("Donor_ID" = md$Donor_ID, "Timepoint" = md$Timepoint, "Infection" = md$`Infection(TB/NI)`, "Batch" = md$Run)

#count number of samples, regardless of run
tp_three <- count(tp_four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
tp_four$Timepoint <- factor(tp_four$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))

tp_four$Infection <- factor(tp_four$Infection, levels = c("NI", "Mtb"))

tp_four$Batch <- factor(tp_four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
tp_three$n <- factor(tp_three$n)

tp_four
tp_three
```
```{r}
p <- ggplot() +
  geom_tile(data = tp_four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = tp_three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_npg(alpha = 0.5)
p
```
Remove the bad data and repeat:

For Run1, these are: 
kallisto_6_AF145_T0_NI_run1
kallisto_6_AF145_T0_Mtb_MOI_5_run1
kallisto_6_AF145_T12_NI_run1
kallisto_6_AF145_T12_Mtb_MOI_5_run1
kallisto_6_AF145_T18_NI_run1
kallisto_6_AF145_T18_Mtb_MOI_5_run1
kallisto_6_AF145_T24_NI_run1
kallisto_6_AF145_T24_Mtb_MOI_5_run1
kallisto_6_AF145_T48_NI_run1
kallisto_6_AF145_T48_Mtb_MOI_5_run1
kallisto_8_AF183_T0_Mtb_MOI_5_run1
kallisto_9_AF193_T0_Mtb_MOI_5_run1
kallisto_3_AF55_T0_Mtb_MOI_5_run1
kallisto_4_AF69_T0_Mtb_MOI_5_run1
kallisto_5_AF95_T0_Mtb_MOI_5_run1
kallisto_5_AF95_T18_NI_run1
kallisto_17_EU02_T0_Mtb_MOI_5_run1
kallisto_18_EU03_T0_Mtb_MOI_5_run1
kallisto_10_EU118_T0_Mtb_MOI_5_run1
kallisto_12_EU122_T0_Mtb_MOI_5_run1
kallisto_13_EU140_T0_Mtb_MOI_5_run1
kallisto_14_EU144_T0_Mtb_MOI_5_run1
kallisto_15_EU148_T0_Mtb_MOI_5_run1
kallisto_16_EU238_T0_Mtb_MOI_5_run1

For Run2, these are: 
14_AF69_T5_NI_S118_L002_R1_001
1_AF31_T3_NI_S41_L001_R1_001
1_AF31_T5_NI_S80_L001_R1_001

Note: Run 2 was already removed

```{r}
md_clean <- filter(md, !filename %in% c("kallisto_6_AF145_T0_NI_run1","kallisto_6_AF145_T0_Mtb_MOI_5_run1","kallisto_6_AF145_T12_NI_run1",
"kallisto_6_AF145_T12_Mtb_MOI_5_run1",
"kallisto_6_AF145_T18_NI_run1",
"kallisto_6_AF145_T18_Mtb_MOI_5_run1",
"kallisto_6_AF145_T24_NI_run1",
"kallisto_6_AF145_T24_Mtb_MOI_5_run1",
"kallisto_6_AF145_T48_NI_run1",
"kallisto_6_AF145_T48_Mtb_MOI_5_run1",
"kallisto_8_AF183_T0_Mtb_MOI_5_run1",
"kallisto_9_AF193_T0_Mtb_MOI_5_run1",
"kallisto_3_AF55_T0_Mtb_MOI_5_run1",
"kallisto_4_AF69_T0_Mtb_MOI_5_run1",
"kallisto_5_AF95_T0_Mtb_MOI_5_run1",
"kallisto_5_AF95_T18_NI_run1",
"kallisto_17_EU02_T0_Mtb_MOI_5_run1",
"kallisto_18_EU03_T0_Mtb_MOI_5_run1",
"kallisto_10_EU118_T0_Mtb_MOI_5_run1",
"kallisto_12_EU122_T0_Mtb_MOI_5_run1",
"kallisto_13_EU140_T0_Mtb_MOI_5_run1",
"kallisto_14_EU144_T0_Mtb_MOI_5_run1",
"kallisto_15_EU148_T0_Mtb_MOI_5_run1",
"kallisto_16_EU238_T0_Mtb_MOI_5_run1"))

md_clean <- filter(md_clean, !filename %in% c("kallisto_14_AF69_T5_NI_run2", "kallisto_1_AF31_T3_NI_run2", "kallisto_1_AF31_T5_NI_run2"))
```

```{r}
#make matrix of values to plot:
four <- data.frame("Donor_ID" = md_clean$Donor_ID, "Timepoint" = md_clean$Timepoint, 
                   "Infection" = md_clean$`Infection(TB/NI)`, "Batch" = md_clean$Run)

#count number of samples, regardless of run
three <- count(four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
four$Timepoint <- factor(four$Timepoint, levels = rev(c("T0","T1","T2","T3","T4","T5","T6","T7","T8","T9",
                                                                       "T10","T12","T14","T16","T18","T24","T30","T36","T42","T48")))

four$Infection <- factor(four$Infection, levels = c("NI", "Mtb"))

four$Batch <- factor(four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
three$n <- factor(three$n)

four
three
```
```{r}
p <- ggplot() +
  geom_tile(data = four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_nejm(alpha = 0.5)
p
```
This still includes the samples that were resequenced, and is therefore a bit confounding. Remove those, showing just run3 for run2+run3 samples. 

Actually, these are already removed. 

```{r}
md_cleaner <- filter(md_clean, !filename %in% c("kallisto_2_AF53_T8_Mtb_run2", "kallisto_6_AF661_T1_NI_run2",
                                                "kallisto_6_AF661_T5_NI_run2", "kallisto_7_AF781_T8_NI_run2", 
                                                "kallisto_8_AF854_T24_Mtb_run2", "kallisto_8_AF854_T3_Mtb_run2", 
                                                "kallisto_9_EU126_T24_Mtb_run2"))
nrow(md_cleaner)
```
### Post-Luis Conversation

After discussing with Luis, we are going to only use samples with timepoints for all individuals. This means droping the following timepoints: T2, T6, T7, T9, T10, T14, T16, T42. We are also dropping T48 because we don't remember why 48 was important. 

Remove these samples and remake the metadata heatmap. 

```{r}
md_cleanest <- filter(md_cleaner, !Timepoint %in% c("T2", "T6", "T7", "T9", "T10", "T14", "T16", "T42", "T48"))
```

```{r}
#make matrix of values to plot:
four <- data.frame("Donor_ID" = md_cleanest$Donor_ID, "Timepoint" = md_cleanest$Timepoint, 
                   "Infection" = md_cleanest$`Infection(TB/NI)`, "Batch" = md_cleanest$Run)

#count number of samples, regardless of run
three <- count(four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
four$Timepoint <- factor(four$Timepoint, levels = rev(c("T0","T1","T3", "T4", "T5","T8","T12","T18","T24","T30","T36")))

four$Infection <- factor(four$Infection, levels = c("NI", "Mtb"))

four$Batch <- factor(four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
three$n <- factor(three$n)

four
three
```
```{r}
p <- ggplot() +
  geom_tile(data = four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_nejm(alpha = 0.5)
p
```
Also drop run4 for Mtb - AF781 - T8 and Mtb - AF53 - T3 and NI - AF278 - T24. Also remove mtb T4. 

```{r}
md_cleanest <- filter(md_cleanest, !Timepoint == "T8" | 
           !`Infection(TB/NI)` == "Mtb" |
           !Donor_ID == "AF781" |
           !Run == "run4") %>%
  filter(!Timepoint == "T3" | 
           !`Infection(TB/NI)` == "Mtb" |
           !Donor_ID == "AF53" |
           !Run == "run4") %>%
  filter(!Timepoint == "T24" | 
           !`Infection(TB/NI)` == "NI" |
           !Donor_ID == "AF278" |
           !Run == "run4") %>%
  filter(!Timepoint == "T4" | 
           !`Infection(TB/NI)` == "Mtb")
```

```{r}
#make matrix of values to plot:
four <- data.frame("Donor_ID" = md_cleanest$Donor_ID, "Timepoint" = md_cleanest$Timepoint, 
                   "Infection" = md_cleanest$`Infection(TB/NI)`, "Batch" = md_cleanest$Run)

#count number of samples, regardless of run
three <- count(four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
four$Timepoint <- factor(four$Timepoint, levels = rev(c("T0","T1","T3", "T4", "T5","T8","T12","T18","T24","T30","T36")))

four$Infection <- factor(four$Infection, levels = c("NI", "Mtb"))

four$Batch <- factor(four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
three$n <- factor(three$n)

four
three
```
```{r}
p <- ggplot() +
  geom_tile(data = four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_aaas(alpha = 0.5)
p
```

### Reassign NI

So NI has different timepoint annotations for the initial timepoints between run1 and run2. In theory they are the same, so we want to reassign T0 to T1 and T4 to both T5 and T3.  

```{r}
#T0 -> T1
md_cleanest <- mutate(md_cleanest, Timepoint = if_else(`Infection(TB/NI)` == "NI" & Run == "run1" & Timepoint == "T0", "T1", Timepoint))

#T4 -> T5
md_cleanest <- mutate(md_cleanest, Timepoint = if_else(`Infection(TB/NI)` == "NI" & Run == "run1" & Timepoint == "T4", "T5", Timepoint))

#duplicate T5 and reassign to T3. Also need to adjust the filename. 
md_cleanest <- md_cleanest %>% 
  bind_rows(
    filter(., `Infection(TB/NI)` == "NI" & Run == "run1" & Timepoint == "T5") %>%
      mutate(Timepoint = "T3") %>% mutate(filename=paste0(filename, "_duplicated"))
  )
```

Okay I think this worked. Now plot. 

```{r}
#make matrix of values to plot:
four <- data.frame("Donor_ID" = md_cleanest$Donor_ID, "Timepoint" = md_cleanest$Timepoint, 
                   "Infection" = md_cleanest$`Infection(TB/NI)`, "Batch" = md_cleanest$Run)

#count number of samples, regardless of run
three <- count(four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
four$Timepoint <- factor(four$Timepoint, levels = rev(c("T0","T1","T3", "T4", "T5","T8","T12","T18","T24","T30","T36")))

four$Infection <- factor(four$Infection, levels = c("NI", "Mtb"))

four$Batch <- factor(four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
three$n <- factor(three$n)

four
three

p <- ggplot() +
  geom_tile(data = four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_aaas(alpha = 0.5)
p
```
Sweet this worked. We are still missing AF95 T18 and EU144 T1 and EU262 T36 in the non-infection. Just use the nearest Timepoint and duplicate. This is okay, especially since it is non-infected. 

Grab T12 for AF95 T18, T3 for EU144 T1, and T30 for EU262 T36. 

```{r}
#T12 for AF95 T18
md_cleanest <- md_cleanest %>% 
  bind_rows(
    filter(., `Infection(TB/NI)` == "NI" & Donor_ID == "AF95" & Timepoint == "T12") %>%
      mutate(Timepoint = "T18") %>% mutate(filename=paste0(filename, "_duplicated"))
  )

#T3 for EU144 T1
md_cleanest <- md_cleanest %>% 
  bind_rows(
    filter(., `Infection(TB/NI)` == "NI" & Donor_ID == "EU144" & Timepoint == "T3") %>%
      mutate(Timepoint = "T1") %>% mutate(filename=paste0(filename, "_duplicated"))
  )

#T30 for EU262 T36
md_cleanest <- md_cleanest %>% 
  bind_rows(
    filter(., `Infection(TB/NI)` == "NI" & Donor_ID == "EU262" & Timepoint == "T30") %>%
      mutate(Timepoint = "T36") %>% mutate(filename=paste0(filename, "_duplicated"))
  )
```

Now plot. 

```{r}
#make matrix of values to plot:
four <- data.frame("Donor_ID" = md_cleanest$Donor_ID, "Timepoint" = md_cleanest$Timepoint, 
                   "Infection" = md_cleanest$`Infection(TB/NI)`, "Batch" = md_cleanest$Run)

#count number of samples, regardless of run
three <- count(four, Donor_ID, Timepoint, Infection)

#adjust factor levels so timepoints and Infection are in order (donors are pretty good already):
four$Timepoint <- factor(four$Timepoint, levels = rev(c("T0","T1","T3", "T4", "T5","T8","T12","T18","T24","T30","T36")))

four$Infection <- factor(four$Infection, levels = c("NI", "Mtb"))

four$Batch <- factor(four$Batch, levels = c("run1", "run2", "run3", "run4"))

# Convert n to a factor
three$n <- factor(three$n)

four
three

p <- ggplot() +
  geom_tile(data = four, aes(x=Donor_ID, y=Timepoint, fill = Batch), color = "white", size=0.2) +
  geom_text(data = three, aes(x=Donor_ID, y=Timepoint, label = n), size = 3) +
  facet_wrap(~Infection) +
  labs(fill = "Count") +
  theme_light() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5), panel.grid = element_blank(), aspect.ratio = 0.75, axis.title.y = element_blank(), 
        axis.title.x = element_blank(), text = element_text(family = "Calibri", colour = "black")) +
  ggsci::scale_fill_aaas(alpha = 0.5)
p
```

```{r}
#write metadata and save pdf
write_tsv(file = "results/metadata_filtered-samples.tsv", x = md_cleanest, col_names = T)
ggsave(plot = p, filename = "results/metadata_filtered-samples_heatmap.pdf", device = cairo_pdf)
```
